Ornaments as indicators of social changes in northeastern Taiwan before and after the European colonial period

### 2020-07-09 analysis/paper/paper.Rmd
Capitalize place name

Despite the relatively short period of the European colonization in Taiwan (38-years-long Dutch occupation and 16-years-long Spanish occupation), the indirect impacts of the Europeans through trade network could go beyond the temporal boundaries and last longer. This is because the use of foreign goods in many Indigenous societies often go through the processes of negotiations, resistance, or transformation that a static time frame may not capture the processes well [@Scheiber2012]. Also, we acknowledged that a fine-grained identification of decades of the European presence from the archaeological record is challenging for this study based on three reasons. The current radiocarbon ages that are unable to provide a high resolution of time breaks, the circulation of foreign goods might start earlier before the official European presence in Taiwan, and the use of the foreign goods in Indigenous societies could continue after the end of the European colonization [@Mitchell2010]. However, those limitations can be minimized with pericolonial approaches that emphasize the wide-ranging effects of the colonial process, allowing discussion of indirect impacts on Indigenous societies with a more practical and dynamic framework not necessarily defined by a fixed colonial period with a start and an end time [@Trabert2018; @Scheiber2012].

We acknowledged the approach to study a continuum of changes of archaeological records over time to avoid the pitfall of temporal divides [@Scheiber2012]; however, we also admitted the convenience of using separated time periods for quantitative comparison between temporal groups of imported ornaments. To mitigate the conflict, we separated the ornaments based on the archaeological context that can reflect the acceptance of diagnostic imported goods by their relatively high frequencies as an indicator of indirect colonial impacts. That is, we separated the time periods to before and after a certain foreign presence using the complete archaeological context that contains higher frequencies of time indicators instead of their first appearance in a given layer. This method is reasonable because it allows not only the flexibility of imported goods entering Indigenous societies depending on the agency of local people, but also the application of quantitative analysis to temporal and spatial changes of imported ornaments.

@@

Based on the copies of original excavation records and fieldwork notes, we have made refinements of the original chronology to assign contexts into the pre-European, European, and the Chinese periods with six steps of assigning and evaluations. First, we determined the artificial layer (mostly 10 cm) with indirect colonial impacts according to the finding of high frequency of time indicators representing the European and the Chinese period. Second, we identified the complete context (normally consists of 2-3 layers) where the layer belongs to based on the description of soil color, soil texture, and unearthed artifacts using original strategrphic data. Third, we checked if there are radiocarbon ages associated with the context as a cross validation. Since we noted that the radiocarbon ages normally represent long time ranges, the ages are not considered if they cannot provide useful information. Fourth, we examined the original fieldwork notes to identify any post-depositional issues that might affect the deposition of time indicators to ensure an appropriate assignment. Fifth, we cross evaluated all adjacent squares to confirm a consistent, and continuous context between them. Sixth, we compared our chronology with six analytic units classified by previous studies to make a final decision [@Hsieh2009; @Wang2011]. More detailed data information is in Supplementary Online Materials. After assigning the contexts into the pre-European contact, the European presence, and the Chinese presence, we decided the chronology of ornaments by examining in which context they were found. Through those assigning steps with validation rules, we believe our chronology is reasonable and suitable for discussing the indirect colonial impacts led by trade networks.

@@

We focus on ornaments from the habitation contexts (Figure \@ref(fig:plot-all-ornaments-count), Table \@ref(tab:tidy-data-for-table)) because these give us the greatest spatial and temporal representation across the three time periods, and so are most informative of social inequality as indicated by uneven distributions of ornaments. The burials are excluded because most burials from the sampling area date to the European period (n = 21), limiting the usefulness of comparisons between the periods. A possible limitation to our chronological resolution is that ornaments could be heirlooms inherited over multiple generations and well-preserved for a long time. This is difficult to rule out completely, but we consider that because there is no continuous increase in ornament frequency over time, we conclude that accumulation and discard of ornaments is not constant. Thus, we assume that changes in the abundance of ornaments reflect relatively continuous discard behaviours rather than accumulations due to collecting of heirlooms.

@@

Although we are not certain of the specific origin of all kinds of beads, research suggest that glass beads and metal ornaments have similar production techniques and composition to those found in China, while gold-foil beads may relate to the economic activities of the Spanish in the South China Sea region [@Chen2011; @Wang2018]. There seem to be no obvious changes in the sources of glass beads or metal ornaments at different phases in the upper component of Kiwulan (1350-1850 AD). However, the glass beads from the lower component (650-1150 AD) demonstrate the composition of m-Na-Al glass and soda plant ash glass that are subtypes of the Indo-Pacific beads, frequently seen in Southeast Asia[@Carter2016; @Francis2002; @Wang2014]. A transition in the origin of glass beads from mainly Southeast Asia to multiple regions including China might indicate different trading networks. A regional network in South China Sea in the earlier period and later involved in a larger scale trading network stimulated by the Europeans and the Chinese [@Hung2016; @Wang2007].

@@

An indirect colonial influence may be indicated at Kiwulan by the greater diversity of ornament types and materials during the European period. Yilan was involved in complex trading networks both on a regional scale with other Indigenous groups and Chinese merchants, and at a global scale with Europeans, including the Dutch and the Spanish. Those trade ornaments have multiple origins, including Southeast Asia and China, and were first introduced into northeastern Taiwan by Chinese merchants before 17th century. Later, trade activities became more frequent and intense in the 17th century due to European activities. The greater diversity and quantity of ornaments likely resulted from participation in large scale exchange networks that stimulated the circulation of different ornament classes. The frequency of overall ornaments and each subtype declines significantly after European influence fades during the Chinese period in the early 19th century. This may be due to a smaller scale of trading networks, the overall decline of Indigenous populations in Yilan, or the adoption of Han Chinese practices. The decline of the population at Kiwulan may be related to the movement of many Indigenous people southwards to Hualien due to the increasing numbers of Han Chinese immigrants who took over their lands at the end of the 18th century [@Chen2007]. Houses and burials may also be a useful source of evidence to understand population size but a proper treatment of those is beyond the scope of the paper.

Archaeological contexts show that ornaments are especially abundant in burial contexts serving as grave goods [@Chen2007]. This supports the interpretation of ornaments as valuable objects functioning as status indicators. Spatial patterns of ornaments in dwelling contexts show that their distribution was clustered during the pre-European and European periods. These clusters are non-random, and are most highly concentrated during the European period. This may indicate that a degree of social inequality based on the uneven distribution of ornaments was already present before European contact, and then it was reinforced and amplified during the European period. A further indicator of increased social inequality is a burial dated to the 17th century that included 60 gold-foil beads, well above the average of 2-3 pieces in the pre-European period [@Chen2007; @Cheng2008]. Based on the finding of that burial, @Cheng2008 proposed a more hierarchical structure of Kiwulan. However, @Hsieh2012's research on burial data suggested a more egalitarian society since a few burials with rich goods are elders that indicates accumulated wealth over time instead of inherited status. Despite the data they used for making inferences are all based on burial contexts, the inferences about Kiwulan social structure are opposite. Our results can provide an additional insight by focusing on the residential area that demonstrated an uneven distribution of ornaments during the European presence, indicating an increasing social inequality. Nevertheless, we still need to explore and compare with other archaeological records to make a better inference to social structure of Kiwulan.

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Ames2010; @Bowles2010; @Drennan2010; @Feinman2000]. The corporate/network model proposed expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power [@Feinman2000]. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labour tasks [@Feinman2000; @Siegel1999].

@@

That said, the evidence from Kiwulan may be consistent with a variety of scenarios of indigenous-colonial relations. The increasing number and concentrated spatial patterns of ornaments may also indicate a practice of cultural resistance against the European intrusion. Resistance to European economic and political demands may be inferred if ornaments were used as a display of social identity and to emphasize the local customs that had existed before European contact [cf. @Rubertone2000]. We know the custom of wearing ornaments in Indigenous societies for attending ceremonial events based on the Chinese historical records in the 19th century [@Chen1963; @Ke1993]. Also, an ethnographic photo photographed by @Mackey1895, pp. 306-307 presents a weaving practice by an Indigenous woman at Yilan, who wore beaded necklace and headpiece. This indicates displaying ornaments is part of their culture, and may be also used in their daily live according to the photo and the archaeological context where ornaments commonly found in the residential area. This custom might be reinforced during the foreign presence as a form of resistance, but we still need more specific evidence from Kiwulan to prefer resistance as the primary mechanism behind the distribution of ornaments over colonial influence. Another scenario is that ornaments were treated as heirlooms, such as carnelian beads and gold-foil beads, that passed from one generation to the next, accumulating at Kiwulan over time. This process would result in a natural increase in ornaments over time, unrelated to colonial influences. As noted above, this pattern is not a good fit for the ornament distribution at Kiwulan, and adds little value in explaining the shifts in spatial patterns.

@@

Analyses of the archaeological record at the peripheries of colonial activity offers a opportunity to understand pervasive but not necessarily dominant European colonial processes at remote Indigenous groups [@Trabert2018]. Similar impacts can be also seen in modern societies today when adopting western products [@AlGhanim2017; @Zheng2015]. Kiwulan in northeastern Taiwan is an exceptional case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan provides valuable insights into the discussion of indirect colonial influence on local societies living beyond the reach of direct European colonial occupation. The frequency and spatial distribution of personal ornaments at Kiwulan present three distinct patterns during different dominant culture interaction periods. The greater amount and diversity of ornament types during the European period reflects an increasing use in ornaments in a colonial context. Before European contact, ornaments were traded into local Indigenous societies via the regional exchange network with Chinese merchants, and viewed as prestige goods in the local Indigenous culture. After the arrival of the Europeans, the exotic and powerful image carried by those ornaments may have intensified, further signaling wealth and privileged trading connections among the inhabitants of Kiwulan. This may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality. This might also indicate an act of intentional resistance to the intrusion of the Europeans by using more ornaments that are symbolic of cultural tradition, but additional evidence is required to confirm this.

>>>>

Despite the relatively short period of the European colonization in Taiwan (38-years-long Dutch occupation and 16-years-long Spanish occupation), the indirect impacts of the Europeans through trade networks could go beyond the temporal boundaries and last longer. This is because the use of foreign goods in many Indigenous societies often goes through the processes of negotiations, resistance, or transformation that a static time frame may not capture the processes well [@Scheiber2012]. Also, we acknowledged that a fine-grained identification of decades of the European presence from the archaeological record is challenging for this study based on three reasons. The current radiocarbon ages that are unable to provide a high resolution of time breaks, the circulation of foreign goods might start earlier before the official European presence in Taiwan, and the use of the foreign goods in Indigenous societies could continue after the end of the European colonization [@Mitchell2010]. However, those limitations can be minimized with pericolonial approaches that emphasize the wide-ranging effects of the colonial process, allowing discussion of indirect impacts on Indigenous societies with a more practical and dynamic framework not necessarily defined by a fixed colonial period with a start and an end time [@Trabert2018; @Scheiber2012].

We acknowledged the approach to study a continuum of changes of archaeological records over time to avoid the pitfall of temporal divides [@Scheiber2012]; however, we also admitted the convenience of using separated time periods for quantitative comparison between temporal groups of imported ornaments. To mitigate the conflict, we separated the ornaments based on the archaeological context that can reflect the acceptance of diagnostic imported goods by their relatively high frequencies as an indicator of indirect colonial impacts. That is, we separated the time periods into a certain period with foreign presence using the complete archaeological context that contains higher frequencies of time indicators instead of the first appearance of those indicators in a given layer. This method is reasonable because it allows a flexibility of imported goods entering Indigenous societies depending on the agency of local people, and a time framework for quantitative analysis for temporal and spatial changes of imported ornaments.

@@

Based on the copies of original excavation records and fieldwork notes, we have made refinements of the original chronology to assign contexts into the pre-European, European, and the Chinese periods with six steps of assigning and evaluations. First, for each unit, we determined the artificial layer (mostly 10 cm) with indirect colonial impacts in terms of the higher frequency of time indicators representing the European and the Chinese period. Second, we identified the archaeological context (normally consists of 2-3 layers) where the layer belongs to based on the description of soil color, soil texture, and density of potsherds and charcoals. We took into account potsherds and charcoals based on the assumption that a larger population can be reflected by the dense distribution of potsherds and charcoals, and Kiwulan was the biggest settlement with a larger population in 1950 [@Nakamura1938, pp. 12]. Third, we checked if there are radiocarbon ages associated with the context as a cross-validation. Since some radiocarbon ages represent long time ranges, we did not consider the ages that cannot provide useful information. Fourth, we explored any post-depositional issues that might affect the deposition of time indicators to ensure an appropriate assignment based on the fieldwork notes and excavation report. Fifth, we cross evaluated all adjacent squares to confirm a consistent, and continuous context between them. Sixth, we compared our chronology with six analytic units classified by previous studies to make a final decision [@Hsieh2009; @Wang2011]. More detailed data information is provided in Supplementary Online Materials. After assigning the contexts into the pre-European contact, the European presence, and the Chinese presence, we decided the chronology of ornaments by examining in which context they were found. Through those assigning steps with validation rules, we believe our chronology is reasonable and suitable for discussing the indirect colonial impacts led by trade networks.

@@

We focus on ornaments from the habitation contexts (Figure \@ref(fig:plot-all-ornaments-count), Table \@ref(tab:tidy-data-for-table)) because these give us the greatest spatial and temporal representation across the three time periods, and so are most informative of social inequality as indicated by uneven distributions of ornaments. The burials are excluded because most burials from the sampling area date to the European period (n = 21), limiting the usefulness of comparisons between the periods. A possible limitation to our chronological resolution is that ornaments could be heirlooms inherited over multiple generations and well-preserved for a long time. This is difficult to rule out completely, but we consider that because there is no continuous increase in ornament frequency over time, we conclude that accumulation and discard of ornaments is not constant. Thus, we assume that changes in the abundance of ornaments reflect relatively continuous discard behaviors rather than accumulations due to collecting of heirlooms.

@@

Although we are not certain of the specific origin of all kinds of beads, research suggests that glass beads and metal ornaments have similar production techniques and composition to those found in China, while gold-foil beads may relate to the economic activities of the Spanish in the South China Sea region [@Chen2011; @Wang2018]. There seem to be no obvious changes in the sources of glass beads or metal ornaments at different phases in the upper component of Kiwulan (1350-1850 AD). However, the glass beads from the lower component (650-1150 AD) demonstrate the composition of m-Na-Al glass and soda plant ash glass that are subtypes of the Indo-Pacific beads, frequently seen in Southeast Asia[@Carter2016; @Francis2002; @Wang2014]. A transition in the origin of glass beads from mainly Southeast Asia to multiple regions including China might indicate different trading networks. A regional network in the South China Sea in the earlier period and later involved in a larger scale trading network stimulated by the Europeans and the Chinese [@Hung2016; @Wang2007].

@@

An indirect colonial influence may be indicated at Kiwulan by the greater diversity of ornament types and materials during the European period. Yilan was involved in complex trading networks both on a regional scale with other Indigenous groups and Chinese merchants, and at a global scale with Europeans, including the Dutch and the Spanish. Those trade ornaments have multiple origins, including Southeast Asia and China, and were first introduced into northeastern Taiwan by Chinese merchants before the 17th century. Later, trade activities became more frequent and intense in the 17th century due to European activities. The greater diversity and quantity of ornaments likely resulted from participation in large scale exchange networks that stimulated the circulation of different ornament classes. The frequency of overall ornaments and each subtype declines significantly after European influence fades during the Chinese period in the early 19th century. This may be due to a smaller scale of trading networks, the overall decline of Indigenous populations in Yilan, or the adoption of Han Chinese practices. The decline of the population at Kiwulan may be related to the movement of many Indigenous people southwards to Hualien due to the increasing numbers of Han Chinese immigrants who took over their lands at the end of the 18th century [@Chen2007]. Houses and burials may also be a useful source of evidence to understand population size but a proper treatment of those is beyond the scope of the paper

Archaeological contexts show that ornaments are especially abundant in burial contexts serving as grave goods [@Chen2007]. This supports the interpretation of ornaments as valuable objects functioning as status indicators. Spatial patterns of ornaments in dwelling contexts show that their distribution was clustered during the pre-European and European periods. These clusters are non-random, and are most highly concentrated during the European period. This may indicate that a degree of social inequality based on the uneven distribution of ornaments was already present before European contact, and then it was reinforced and amplified during the European period. A further indicator of increased social inequality is a burial dated to the 17th century that included 60 gold-foil beads, well above the average of 2-3 pieces in the pre-European period [@Chen2007; @Cheng2008]. Based on the finding of that burial, @Cheng2008 proposed a more hierarchical structure of Kiwulan. However, @Hsieh2012's research on burial data suggested a more egalitarian society since a few burials with rich goods are elders that indicate accumulated wealth over time instead of inherited status. Despite the data they used for making inferences are all based on burial contexts, the inferences about Kiwulan social structure are the opposite. Our results can provide an additional insight by focusing on the residential area that demonstrated an uneven distribution of ornaments during the European presence, indicating an increasing social inequality. Nevertheless, we still need to explore and compare with other archaeological records to make a better inference to the social structure of Kiwulan.

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Ames2010; @Bowles2010; @Drennan2010; @Feinman2000]. The corporate/network model proposed expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power [@Feinman2000]. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labor tasks [@Feinman2000; @Siegel1999].

@@

That said, the evidence from Kiwulan may be consistent with a variety of scenarios of indigenous-colonial relations. The increasing number and concentrated spatial patterns of ornaments may also indicate a practice of cultural resistance against the European intrusion. Resistance to European economic and political demands may be inferred if ornaments were used as a display of social identity and to emphasize the local customs that had existed before European contact [cf. @Rubertone2000]. We know the custom of wearing ornaments in Indigenous societies for attending ceremonial events based on the Chinese historical records in the 19th century [@Chen1963; @Ke1993]. Also, an ethnographic photo photographed by @Mackey1895, pp. 306-307 presents a weaving practice by an Indigenous woman at Yilan, who wore beaded necklace and headpiece. This indicates displaying ornaments is part of their culture, and likely to be also used in their daily life according to the photo and the archaeological context where ornaments commonly found in the residential area. This custom might be reinforced during the foreign presence as a form of resistance, but we still need more specific evidence from Kiwulan to prefer resistance as the primary mechanism behind the distribution of ornaments over colonial influence. Another scenario is that ornaments were treated as heirlooms, such as carnelian beads and gold-foil beads, that passed from one generation to the next, accumulating at Kiwulan over time. This process would result in a natural increase in ornaments over time, unrelated to colonial influences. As noted above, this pattern is not a good fit for the ornament distribution at Kiwulan, and adds little value in explaining the shifts in spatial patterns.

@@

Analyses of the archaeological record at the peripheries of colonial activity offers an opportunity to understand pervasive but not necessarily dominant European colonial processes at remote Indigenous groups [@Trabert2018]. Similar impacts can be also seen in modern societies today when adopting western products [@AlGhanim2017; @Zheng2015]. Kiwulan in northeastern Taiwan is an exceptional case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan provides valuable insights into the discussion of indirect colonial influence on local societies living beyond the reach of direct European colonial occupation. The frequency and spatial distribution of personal ornaments at Kiwulan present three distinct patterns during different dominant culture interaction periods. The greater amount and diversity of ornament types during the European period reflects an increasing use in ornaments in a colonial context. Before European contact, ornaments were traded into local Indigenous societies via the regional exchange network with Chinese merchants, and viewed as prestige goods in the local Indigenous culture. After the arrival of the Europeans, the exotic and powerful image carried by those ornaments may have intensified, further signaling wealth and privileged trading connections among the inhabitants of Kiwulan. This may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality. This might also indicate an act of intentional resistance to the intrusion of the Europeans by using more ornaments that are symbolic of cultural tradition, but additional evidence is required to confirm this.
